Long-term trends in the contribution of major causes of death to the black-white life expectancy gap by US state. A retrospective study of all deaths between 1969 and 2013

### 2017-06-29 Manuscript/Appendix-agestd-mort-rates.Rmd
updated fig ref and appdx move

**Figure S1: Age-standardized trends in CVD-related mortality in black and white men, United States, 1969-2013**

@@

**Figure S2: Age-standardized trends in CVD-related mortality in black and white women, United States, 1969-2013**

@@

**Figure S3: Age-standardized trends in cancer-related mortality in black and white men, United States, 1969-2013**

@@

**Figure S4: Age-standardized trends in cancer-related mortality in black and white women, United States, 1969-2013**

@@

**Figure S5: Age-standardized trends in mortality related to non-communicable disease in black and white men, United States, 1969-2013**

@@

**Figure S6: Age-standardized trends in mortality related to non-communicable disease in black and white women, United States, 1969-2013**

@@

**Figure S7: Age-standardized trends in injury-related mortality in black and white men, United States, 1969-2013**

@@

**Figure S8: Age-standardized trends in injury-related mortality in black and white women, United States, 1969-2013**

@@

**Figure S9: Age-standardized trends in mortality related to communicable disease in black and white men, United States, 1969-2013**

@@

**Figure S10: Age-standardized trends in mortality related to communicable disease in black and white women, United States, 1969-2013**

@@

**Figure S11: Age-standardized trends in mortality from all other causes in black and white men, United States, 1969-2013**

@@

**Figure S12: Age-standardized trends in mortality from all other causes in black and white women, United States, 1969-2013**

>>>>

**Figure S3: Age-standardized trends in CVD-related mortality in black and white men, United States, 1969-2013**

@@

**Figure S4: Age-standardized trends in CVD-related mortality in black and white women, United States, 1969-2013**

@@

**Figure S5: Age-standardized trends in cancer-related mortality in black and white men, United States, 1969-2013**

@@

**Figure S6: Age-standardized trends in cancer-related mortality in black and white women, United States, 1969-2013**

@@

**Figure S7: Age-standardized trends in mortality related to non-communicable disease in black and white men, United States, 1969-2013**

@@

**Figure S8: Age-standardized trends in mortality related to non-communicable disease in black and white women, United States, 1969-2013**

@@

**Figure S9: Age-standardized trends in injury-related mortality in black and white men, United States, 1969-2013**

@@

**Figure S10: Age-standardized trends in injury-related mortality in black and white women, United States, 1969-2013**

@@

**Figure S11: Age-standardized trends in mortality related to communicable disease in black and white men, United States, 1969-2013**

@@

**Figure S12: Age-standardized trends in mortality related to communicable disease in black and white women, United States, 1969-2013**

@@

**Figure S13: Age-standardized trends in mortality from all other causes in black and white men, United States, 1969-2013**

@@

**Figure S14: Age-standardized trends in mortality from all other causes in black and white women, United States, 1969-2013**

### 2017-06-29 Manuscript/Manuscript_COD_Trends_BMJOpen.Rmd
updated fig ref and appdx move

###Abstract (check if 300 words or fewer)

@@

Models were fit using Markov chain Monte Carlo (MCMC). These models were run separately for every combination of state, race, gender, and cause of death. They were fit using two chains of length 10,000, with a 2,000 iteration burn-in period to ensure convergence. One thousand iterations remained after burn-in and sampling every eighth iteration to reduce autocorrelation among the MCMC samples. Each iteration produced death count estimates that were smoothed over time for every cause of death within each age group, which were used to estimate life expectancy, the life expectancy gap, and perform a cause of death decomposition. These 1,000 decompositions form the posterior distribution from which the credible intervals were estimated using the 2.5^th^ and 97.5^th^ percentiles.

@@

We also calculated state-level time trends in age-standardized cause-specific mortality rates separately for each combination of race and gender, to aid the interpretation of the decomposition results. We used the US standard population from the year 2000.^13^

@@

To describe the national trend in each causes' contribution to the life expectancy gap over time, we used random effects meta regression to model each state's mean contribution estimated from the previous step as a function of a state random effect and a cubic spline for year (estimated using four internal knots). The model was inversely-weighted by the standard error of the mean estimate in each state-year. To identify states divergent from the national trend, we compared the state-specific trends estimated in the previous step to the national pattern estimated by setting the random effect to zero.

@@

**Figures 1 and 2** display trends in the life expectancy gap for men and women, respectively. For men and women, the life expectancy gap has decreased since 1969, although several states exhibited a period of increasing inequality that peaked in the early 1990s, especially among men.

**Figures 2 - 14** display trends in the contribution of each broad cause of death to the life expectancy gap for men and women, in years, for each state. The blue line depicts the mean of the smoothed contribution estimated for each state, such that states can have very different trend lines. These lines are overlaid with the meta-estimated national trend, shown with a red line. We supplement these main results with smoothed estimates of the age-standardized mortality rates for each cause of death (**Figures S1-S12**).

@@

Cardiovascular disease (CVD) is the most common cause of death for both men and women of both races. Mortality from CVD decreased markedly since 1969 for all groups (**Figures S1 and S2**), but white men underwent relatively larger mortality decreases than black men. As a result, the contribution of CVD to the black-white life expectancy gap among men increased modestly for most states (**Figure 3**). Black women experienced relatively larger mortality decreases over time compared to white women, resulting in a decrease to the CVD contribution to the life expectancy gap that was greater than one year in several states (**Figure 4**). For both genders, Michigan, Wisconsin, California, and DC demonstrated higher-than-average increases in the contribution of CVD over time, whereas Massachusetts and New York showed persistently lower CVD contributions.

@@

For all groups, mortality from cancer peaked in the mid 1990s and decreased thereafter (**Figures S3 and S4**). Among men, this led to an inverted-U shape pattern in cancer's contribution to the life expectancy gap (**Figure 5**). For women, although cancer-related mortality continues to be higher among blacks in many states, the mortality trends were similar between blacks and whites, resulting in a flat general pattern in the contribution of cancer to the life expectancy over time (**Figure 6**). For both genders, DC and Wisconsin exhibit the largest cancer contributions (and also exhibit high initial values among men), while New York and Massachusetts experienced large decreases in the contribution of cancer to the life expectancy gap.

@@

Mortality from non-communicable disease increased over time for men in most states, with DC as a notable exception (**Figure S5**). Recently, there was a slight decrease in mortality rates, especially among black men. These trends are similar among women, except that black women exhibited a sharper decline in mortality in recent years, eliminating the mortality difference in several states (**Figure S6**). As such, the contribution of non-communicable disease to the life expectancy gap was generally flat among men with a recent decrease, and exhibited a gradual decrease over time among women (**Figures 7 and 8**). As with other causes, New York and Massachusetts exhibited decreasing contributions, and Nevada also made substantial progress. Furthermore, while DC's mortality rates decreased, they presently have the highest contribution of non-communicable disease to the gap.

@@

Trends in mortalilty from injuries differ for blacks and whites. Over time, injury-related mortality has decreased precipitously among blacks, although an epidemic of homicide resulted in heightened mortality from injuries among black men during the 1980s and early 1990s (**Figure S7**).^3^ For whites, most states exhibit an initial period of gradual decreases in injury mortality, followed by a period of gradual increases, especially among white women (**Figure S8**). Thus, injury mortality rates among blacks are equivalent, or lower than those among whites in many states, with white women demonstrating sharper recent increases. The contribution of injuries to the mortality gap has therefore generally decreased over time, and has become negative in those states that now have lower injury-related mortality among blacks than whites (**Figures 9 and 10**). Among men, DC, Illinois, Michigan and Missouri exhibit the largest recent injury-related contributions. Among women, many southern states exhibit negative contributions (indicating lower injury-related mortality among black women), while DC, Illinois, and Michigan having higher contributions than the general trend.

@@

Of the major causes that we considered, mortality from communicable disease was the lowest, although the 1990s exhibited an enormous increase in mortality at the height of the HIV/AIDS epidemic before antiretroviral therapy was available (**Figures S9 and S10**). This spike was greater among men, especially blacks, and among states with larger urban centers. The contribution of communicable disease to the life expectancy gap therefore exhibited a peak in the mid 1990s, with states diverging from this general pattern according to the intensity of their local epidemics (**Figures 11 and 12**). DC still exhibits higher communicable disease-related contributions among men, while New Jersey and Florida additionally show higher contributions among women.

@@

Mortality from all other causes generally decreased earlier in the time period, especially among blacks (**Figures S11 and S12**). White men and women have shown increasing mortality rates from all other causes since the 1980s, along with black women, whereas black men have continued to experience decreasing rates within several states. As such, the contribution of all other causes to the life expectancy gap has generally decreased over time, and is virtually zero in New York, Massachusetts, and Rhode Island for both genders (**Figures 13 and 14**). The most disparate states are Pennsylvania and DC, with much larger contributions estimated for all other causes in the earlier years within Pennsylvania and in the later years within DC.

Because the majority of infant deaths are classified as being due to all other causes in our analysis, we also tried to separate any impact of all other causes from those of infant mortality more generally. We conducted a sensitivity analysis to calculate the contribution of deaths due to all other causes that occurred during the first year of life to the life expectancy gap (Figures 15 and 16). We found that 43% of the contribution of all other causes to the gap was due to differences in infant mortality in 1969. In 2013, this proportion reached 53%. Thus, the trend in the all other causes contribution partially reflects a decreasing contribution from infant mortality.

@@

**Figure 1: State-level trends in the black-white life expectancy gap in men, United States, 1969-2013**

**Figure 1 Legend:** The bands illustrate the 95% credible interval around the mean trend line. States with more imprecise estimates (often due to small black populations) have wider credible intervals. The estimated life expectancy gap in California in 1972 is much higher than in surrounding years. This anomaly is also illustrated in Figure 3, and is a product of an unexpectedly low CVD-related mortality rate among white males in California in 1972, likely denoting a coding error. Thus the data for this state and year should be interpreted with caution.

**Figure 2: State-level trends in the black-white life expectancy gap in women, United States, 1969-2013**

**Figure 3: Smoothed state-level trends in the contribution of CVD to the life expectancy gap vs. the national pattern in men, United States, 1969-2013**

@@

**Figure 3 Legend:** The blue line and light blue band illustrate the mean and 95% credible interval for each state's estimated trend in the contribution of CVD to the black-white life expectancy gap over time. The red line is the same across the panels and illustrates the national pattern in the contribution, to add in the identification of similar and divergent states. The early spike in California's trend line illustrates the unexpectedly high estimate of the CVD contribution to the black-white life expectancy gap among California males in 1972. This anomolous estimate reflects the unexpectedly low CVD-related mortality rate among white males in California in 1972, and likely denotes a coding error. Thus the data for this state and year should be interpreted with caution.

**Figure 4: Smoothed state-level trends in the contribution of CVD to the life expectancy gap vs. the national pattern in women, United States, 1969-2013**

@@

**Figure 5: Smoothed state-level trends in the contribution of cancer disease to the life expectancy gap vs. the national pattern in men, United States, 1969-2013**

@@

**Figure 6: Smoothed state-level trends in the contribution of cancer disease to the life expectancy gap vs. the national pattern in women, United States, 1969-2013**

@@

**Figure 7: Smoothed state-level trends in the contribution of non-communicable disease to the life expectancy gap vs. the national pattern in men, United States, 1969-2013**

@@

**Figure 8: Smoothed state-level trends in the contribution of non-communicable disease to the life expectancy gap vs. the national pattern in women, United States, 1969-2013**

@@

**Figure 9: Smoothed state-level trends in the contribution of injuries to the life expectancy gap vs. the national pattern in men, United States, 1969-2013**

@@

**Figure 10: Smoothed state-level trends in the contribution of injuries to the life expectancy gap vs. the national pattern in women, United States, 1969-2013**

@@

**Figure 11: Smoothed state-level trends in the contribution of communicable disease to the life expectancy gap vs. the national pattern in men, United States, 1969-2013**

@@

**Figure 12: Smoothed state-level trends in the contribution of communicable disease to the life expectancy gap vs. the national pattern in women, United States, 1969-2013**

@@

**Figure 13: Smoothed state-level trends in the contribution of all other causes to the life expectancy gap vs. the national pattern in men, United States, 1969-2013**

@@

**Figure 14: Smoothed state-level trends in the contribution of all other causes to the life expectancy gap vs. the national pattern in women, United States, 1969-2013**

@@

**Figure 15: Smoothed state-level trends in the contribution of all other causes to the life expectancy gap vs. the national pattern in men younger than one, 1969-2013**

**Figure 16: Smoothed state-level trends in the contribution of all other causes to the life expectancy gap vs. the national pattern in women younger than one, 1969-2013**

@@

This work was supported by Dr. Kaufman's Canada Research Chair in Health Disparities. Corinne Riddell is a post-doctoral researcher at McGill University, whose salary was supported by Dr. Kaufman's Canada Research Chair in Health Disparities. Kathryn Morrison is a post-doctoral researcher at McGill University. Sam Harper was supported by a Chercheur Boursier Junior 2 from the Fonds de la Recherche en Santé du Québec. Jay Kaufman was supported by a Canada Research Chair in Health Disparities.

@@

>>>>

###Abstract

@@

Models were fit using Markov chain Monte Carlo (MCMC). These models were run separately for every combination of state, race, gender, and cause of death. They were fit using two chains of length 10,000, with a 2,000 iteration burn-in period to ensure convergence. One thousand iterations remained after burn-in and sampling every eighth iteration to reduce autocorrelation among the MCMC samples. Each iteration produced death count estimates that were smoothed over time for every cause of death within each age group, which were used to estimate life expectancy, the life expectancy gap, and perform a cause of death decomposition. These 1,000 decompositions form the posterior distribution from which the credible intervals were estimated using the 2.5^th^ and 97.5^th^ percentiles. This modeling was done in R (version 3.3.2) using the *rjags* package.

@@

We also calculated state-level trends in age-standardized cause-specific mortality rates separately for each combination of race and gender, to aid the interpretation of the decomposition results. We used the US standard population from the year 2000.^13^ The calculations described in this section were performed using functions written by the first author.

@@

To describe the national trend in each causes' contribution to the life expectancy gap over time, we used random effects meta regression to model each state's mean contribution estimated from the previous step as a function of a state random effect and a cubic spline for year (estimated using four internal knots). The model was inversely-weighted by the standard error of the mean estimate in each state-year. To identify states divergent from the national trend, we compared the state-specific trends estimated in the previous step to the national pattern estimated by setting the random effect to zero. These estimates relied on the R *lme4* package for random effects modeling, and the *splines* package to define cubic splines.

@@

**Figure S1 and S2** display trends in the life expectancy gap for men and women, respectively. For men and women, the life expectancy gap has decreased since 1969, although several states exhibited a period of increasing inequality that peaked in the early 1990s, especially among men.

**Figures 1 - 6** display trends in the contribution of each broad cause of death to the life expectancy gap for men and women, in years, for each state. The blue line depicts the mean of the smoothed contribution estimated for each state, such that states can have very different trend lines. These lines are overlaid with the meta-estimated national trend, shown with a red line. We supplement these main results with smoothed estimates of the age-standardized mortality rates for each cause of death (**Figures S3-S14**).

@@

Cardiovascular disease (CVD) is the most common cause of death for both men and women of both races. Mortality from CVD decreased markedly since 1969 for all groups (**Figures S3 and S4**), but white men underwent relatively larger mortality decreases than black men. As a result, the contribution of CVD to the black-white life expectancy gap among men increased modestly for most states (**Figure 1A**). Black women experienced relatively larger mortality decreases over time compared to white women, resulting in a decrease to the CVD contribution to the life expectancy gap that was greater than one year in several states (**Figure 1B**). For both genders, Michigan, Wisconsin, California, and DC demonstrated higher-than-average increases in the contribution of CVD over time, whereas Massachusetts and New York showed persistently lower CVD contributions.

@@

For all groups, mortality from cancer peaked in the mid 1990s and decreased thereafter (**Figures S5 and S6**). Among men, this led to an inverted-U shape pattern in cancer's contribution to the life expectancy gap (**Figure 2A**). For women, although cancer-related mortality continues to be higher among blacks in many states, the mortality trends were similar between blacks and whites, resulting in a flat general pattern in the contribution of cancer to the life expectancy over time (**Figure 2B**). For both genders, DC and Wisconsin exhibit the largest cancer contributions (and also exhibit high initial values among men), while New York and Massachusetts experienced large decreases in the contribution of cancer to the life expectancy gap.

@@

Mortality from non-communicable disease increased over time for men in most states, with DC as a notable exception (**Figure S7**). Recently, there was a slight decrease in mortality rates, especially among black men. These trends are similar among women, except that black women exhibited a sharper decline in mortality in recent years, eliminating the mortality difference in several states (**Figure S8**). As such, the contribution of non-communicable disease to the life expectancy gap was generally flat among men with a recent decrease, and exhibited a gradual decrease over time among women (**Figures 3A and 3B**). As with other causes, New York and Massachusetts exhibited decreasing contributions, and Nevada also made substantial progress. Furthermore, while DC's mortality rates decreased, they presently have the highest contribution of non-communicable disease to the gap.

@@

Trends in mortalilty from injuries differ for blacks and whites. Over time, injury-related mortality has decreased precipitously among blacks, although an epidemic of homicide resulted in heightened mortality from injuries among black men during the 1980s and early 1990s (**Figure S9**).^3^ For whites, most states exhibit an initial period of gradual decreases in injury mortality, followed by a period of gradual increases, especially among white women (**Figure S10**). Thus, injury mortality rates among blacks are equivalent, or lower than those among whites in many states, with white women demonstrating sharper recent increases. The contribution of injuries to the mortality gap has therefore generally decreased over time, and has become negative in those states that now have lower injury-related mortality among blacks than whites (**Figures 4A and 4B**). Among men, DC, Illinois, Michigan and Missouri exhibit the largest recent injury-related contributions. Among women, many southern states exhibit negative contributions (indicating lower injury-related mortality among black women), while DC, Illinois, and Michigan having higher contributions than the general trend.

@@

Of the major causes that we considered, mortality from communicable disease was the lowest, although the 1990s exhibited an enormous increase in mortality at the height of the HIV/AIDS epidemic before antiretroviral therapy was available (**Figures S11 and S12**). This spike was greater among men, especially blacks, and among states with larger urban centers. The contribution of communicable disease to the life expectancy gap therefore exhibited a peak in the mid 1990s, with states diverging from this general pattern according to the intensity of their local epidemics (**Figures 5A and 5B**). DC still exhibits higher communicable disease-related contributions among men, while New Jersey and Florida additionally show higher contributions among women.

@@

Mortality from all other causes generally decreased earlier in the time period, especially among blacks (**Figures S13 and S14**). White men and women have shown increasing mortality rates from all other causes since the 1980s, along with black women, whereas black men have continued to experience decreasing rates within several states. As such, the contribution of all other causes to the life expectancy gap has generally decreased over time, and is virtually zero in New York, Massachusetts, and Rhode Island for both genders (**Figures 6A and 6B**). The most disparate states are Pennsylvania and DC, with much larger contributions estimated for all other causes in the earlier years within Pennsylvania and in the later years within DC.

Because the majority of infant deaths are classified as being due to all other causes in our analysis, we also tried to separate any impact of all other causes from those of infant mortality more generally. We conducted a sensitivity analysis to calculate the contribution of deaths due to all other causes that occurred during the first year of life to the life expectancy gap (**Figures S15 and S16**). We found that 43% of the contribution of all other causes to the gap was due to differences in infant mortality in 1969. In 2013, this proportion reached 53%. Thus, the trend in the all other causes contribution partially reflects a decreasing contribution from infant mortality.

@@

**Figure 1A: Smoothed state-level trends in the contribution of CVD to the life expectancy gap vs. the national pattern in men, United States, 1969-2013**

@@

**Figure 1 Legend:** The blue line and light blue band illustrate the mean and 95% credible interval for each state's estimated trend in the contribution of CVD to the black-white life expectancy gap over time. The red line is the same across the panels and illustrates the national pattern in the contribution, to add in the identification of similar and divergent states. The early spike in California's trend line illustrates the unexpectedly high estimate of the CVD contribution to the black-white life expectancy gap among California males in 1972. This anomolous estimate reflects the unexpectedly low CVD-related mortality rate among white males in California in 1972, and likely denotes a coding error. Thus the data for this state and year should be interpreted with caution.

**Figure 1B: Smoothed state-level trends in the contribution of CVD to the life expectancy gap vs. the national pattern in women, United States, 1969-2013**

@@

**Figure 2A: Smoothed state-level trends in the contribution of cancer disease to the life expectancy gap vs. the national pattern in men, United States, 1969-2013**

@@

**Figure 2B: Smoothed state-level trends in the contribution of cancer disease to the life expectancy gap vs. the national pattern in women, United States, 1969-2013**

@@

**Figure 3A: Smoothed state-level trends in the contribution of non-communicable disease to the life expectancy gap vs. the national pattern in men, United States, 1969-2013**

@@

**Figure 3B: Smoothed state-level trends in the contribution of non-communicable disease to the life expectancy gap vs. the national pattern in women, United States, 1969-2013**

@@

**Figure 4A: Smoothed state-level trends in the contribution of injuries to the life expectancy gap vs. the national pattern in men, United States, 1969-2013**

@@

**Figure 4B: Smoothed state-level trends in the contribution of injuries to the life expectancy gap vs. the national pattern in women, United States, 1969-2013**

@@

**Figure 5A: Smoothed state-level trends in the contribution of communicable disease to the life expectancy gap vs. the national pattern in men, United States, 1969-2013**

@@

**Figure 5B: Smoothed state-level trends in the contribution of communicable disease to the life expectancy gap vs. the national pattern in women, United States, 1969-2013**

@@

**Figure 6A: Smoothed state-level trends in the contribution of all other causes to the life expectancy gap vs. the national pattern in men, United States, 1969-2013**

@@

**Figure 6B: Smoothed state-level trends in the contribution of all other causes to the life expectancy gap vs. the national pattern in women, United States, 1969-2013**

@@

@@

This research received no specific grant from any funding agency in the public, commercial or not-for-profit sectors.

Corinne Riddell is a post-doctoral researcher at McGill University, whose salary was supported by Dr. Kaufman's Canada Research Chair in Health Disparities. Kathryn Morrison is a post-doctoral researcher at McGill University. Sam Harper was supported by a Chercheur Boursier Junior 2 from the Fonds de la Recherche en Santé du Québec. Jay Kaufman was supported by a Canada Research Chair in Health Disparities.

@@

###Supplementary appendix

**Figure S1: State-level trends in the black-white life expectancy gap in men, United States, 1969-2013**

**Figure S1 Legend:** The bands illustrate the 95% credible interval around the mean trend line. States with more imprecise estimates (often due to small black populations) have wider credible intervals. The estimated life expectancy gap in California in 1972 is much higher than in surrounding years. This anomaly is also illustrated in Figure 3, and is a product of an unexpectedly low CVD-related mortality rate among white males in California in 1972, likely denoting a coding error. Thus the data for this state and year should be interpreted with caution.

**Figure S2: State-level trends in the black-white life expectancy gap in women, United States, 1969-2013**

**Figure S15: Smoothed state-level trends in the contribution of all other causes to the life expectancy gap vs. the national pattern in men younger than one, 1969-2013**

**Figure S16: Smoothed state-level trends in the contribution of all other causes to the life expectancy gap vs. the national pattern in women younger than one, 1969-2013**
